A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-10 paper.tex
More fleshing out of discussion

In this paper we [turn into proper round up paragraph later]
\begin{enumerate}
\item Discuss the classical Ancestral Recombination Graph formulation,
and show how it derives from (and is limited by) the coalescent with
recombination.
\item In the following sections we
show that, although the ARG-as-a-data-structure perspective is
now the most common interpretation, the \emph{form} of that
data structure is still closely linked to the original
stochastic process, and that this form is inflexible
in the types of genetic inheritance that can be described,
computationally inefficient, and
forces complete (and potentially spurious) precision about
the location and timing
of recombination events. % (and coalescent events?)
\item Suggest a definition of ARGs that is free of these limitations,
where the formulation is derived from a (normally diploid) pedigree rather than from
coalescent approximations, and where the genetic
material which is transmitted is described by annotations attached to
edges rather than nodes.
\item Suggest a classification of ARG nodes that helps us to understand
the properties of the structures inferred by different methods, and illustrate
with some examples.
% Perhaps we should also mention something about simplification and unknowable
% nodes, e.g. ``show how some of the nodes are not required for useful interpretation
% of ancestry''.
\item Despite its status as ``the ideal data structure for
population genomic analysis''~\citep{rasmussen2014genome}, there has been
little discussion of its computational properties. We discuss the
relationship between the classical concept of an ARG and the
recently introduced succinct tree sequence data structure.
\end{enumerate}

@@

% JK: This is very rough, just jotting down the basic ideas,
% will need substantial revision.
It is not clear that such complete information will always
(or indeed ever) be available from inferences. We may be
able to precisely identify the location and timing
of recent recombinations (particularly if detailed pedigree
information is available) but beyond this, the information
required simply isn't present. When sampling from a
process such as the Sequentially Markov
Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
like ARGweaver~\citep{rasmussen2014genome}, it is true that
we do obtain a fully resolved ARG with precise
information about recombination, but it is reasonable
to question how much information from the data is used
to support older recombinations, in particular.

@@

@@

\section*{Discussion}

Discussion points (may or may not be used, certainly would be heavily revised)

\begin{itemize}

\item Recap quickly distinction bewtween aARG and gARG.Then:
    This, coupled with
    the explicit association of genome coordinates with edges, allows
    us to represent even the bewilderingly complex patterns of relatedness
    present in modern datasets.

\item In general, discussions tend to contrast an ARG
on one hand as holding \emph{complete} information about recombination
(and therefore between-tree correlation structure)
and a sequence of local trees on the other (with, by implication,
no direct information about between-tree correlations).
A key goal of this article is to show that this is a false dichotomy:
a \emph{continuum} of information about recombination and between-tree
correlation structure can be represented and inferred, and all
such levels of information are useful, even if only for computational
efficiency.

\item This idea of a continuum of information about recombination
being inferrable and representable is consistent with the ideas
discussed by~\cite{shipilina2022origin}. There a haplotype block
is synonymous with coalescent nodes in the ARG, which we can generalise
as nodes in a gARG with at least two children. In principle, these
nodes are identifiable by the mutations that occur on the outbound
edges of the node.
Their definition of an ARG is interesting, as they emphasise that it
is composed of events, but illustate the genomes. [Nice to mention
this connection somehere if we can, see Fig B1]

\item It is unfortunate that we are adding yet another meaning to the
term ARG here, but all we're really doing is making the implicit
assumptions that others have made (i.e., describing the output of
Relate, tsinfer etc as ARGs) and making them concrete. We need to have a
term to describe all of these things, and since people are already using
ARG for this, we might as well make it rigorous. We can then move away
from the unfruitful discussion about which methods infer a ``true'' or
``real'' ARG (none do, really, under the original definitions) and
instead start discussing the \emph{properties} of these ARGs. (See
next point also.)

\item Terminology based around an inferred ancestry being a ``true''
ARG is impoverished and misleading. The only reasonable interpretation
of a ARG being true or not is whether it's a sample from the coalescent.
Most inference methods (and all that scale well) do not. In any case,
the coalescent is just one model, and a highly idealised one. It is
exceedingly unlikely that the actual ARG describing the ancestry
of any biological population is ``true'' ARG in this sense. The terminology is
impoverished because it reduces the possibilities available
to something either being an ARG or not, whereas there are many different
interesting properties of these structures that we should be able to
discuss and compare.

\item Comparison metrics are a problem. All evaluations has essentially
looked at either some form of pairwise comparison of trees along the genome
using tree metrics as proposed by \citet{kuhner2015assessing},
or looked at the distributions of TMRCAs. No
one has succeeded in performing any comparisons of ARGs beyond
regarding them as a sequence of independent trees. If we genuinely believe
that ARGs are more than this, then we need to devise some metrics that
consider more than the local tree structure.

\end{itemize}

>>>>

In this paper we [turn into proper round up paragraph later]
\begin{enumerate}
\item Discuss the classical Ancestral Recombination Graph formulation,
and show how it derives from (and is limited by) the coalescent with
recombination.
\item In the following sections we
show that, although the ARG-as-a-data-structure perspective is
now the most common interpretation, the \emph{form} of that
data structure is still closely linked to the original
stochastic process, and that this form is inflexible
in the types of genetic inheritance that can be described,
computationally inefficient, and
forces complete (and potentially spurious) precision about
the location and timing
of recombination events. % (and coalescent events?)
\item Suggest a definition of ARGs that is free of these limitations,
where the formulation is derived from a (normally diploid) pedigree rather than from
coalescent approximations, and where the genetic
material which is transmitted is described by annotations attached to
edges rather than nodes.
\item Suggest a classification of ARG nodes that helps us to understand
the properties of the structures inferred by different methods, and illustrate
with some examples.
% Perhaps we should also mention something about simplification and unknowable
% nodes, e.g. ``show how some of the nodes are not required for useful interpretation
% of ancestry''.
\item Despite its status as ``the ideal data structure for
population genomic analysis''~\citep{rasmussen2014genome}, there has been
little discussion of its computational properties. We discuss the
relationship between the classical concept of an ARG and the
recently introduced succinct tree sequence data structure.

\item This, coupled with
    the explicit association of genome coordinates with edges, allows
    us to represent even the bewilderingly complex patterns of relatedness
    present in modern datasets.

\end{enumerate}

@@

% JK: This is very rough, just jotting down the basic ideas,
% will need substantial revision.
It is not clear that such complete information will always
(or indeed ever) be available from inferences~\cite{shipilina2022origin}.
 We may be
able to precisely identify the location and timing
of recent recombinations (particularly if detailed pedigree
information is available) but beyond this, the information
required simply isn't present. When sampling from a
process such as the Sequentially Markov
Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
like ARGweaver~\citep{rasmussen2014genome}, it is true that
we do obtain a fully resolved ARG with precise
information about recombination, but it is reasonable
to question how much information from the data is used
to support older recombinations, in particular.

% This idea of a continuum of information about recombination
% being inferrable and representable is consistent with the ideas
% discussed by~\cite{shipilina2022origin}. There a haplotype block
% is synonymous with coalescent nodes in the ARG, which we can generalise
% as nodes in a gARG with at least two children. In principle, these
% nodes are identifiable by the mutations that occur on the outbound
% edges of the node.

@@

In general, discussions tend to contrast an ARG
on one hand as holding \emph{complete} information about recombination
(and therefore between-tree correlation structure)
and a sequence of local trees on the other (with, by implication,
no direct information about between-tree correlations).
This is a false dichotomy:
a \emph{continuum} of information about recombination and between-tree
correlation structure can be represented and inferred, and all
such levels of information are useful, even if only for computational
efficiency.

@@

\section*{Discussion}
Recent breakthroughs have made ARG inference methods practical
for the first time, and there has been a surge of interest
in inference methods, their evaluation and application.
The prospect of ARGs being used routinely in population
and statistical genetics applications is tantalising,
but in reality there is  substantial work to be done to
enable this, and much remains to be done.

A necessary first step towards this goal is some terminological
clarity. As discussed in Appendix XXX, the term ARG has several
subtly different interpretations depending on context. The
trend to decouple the ARG from its origins as a stochastic
process and instead use the term as a more general representation of any
recombinant genetic ancestry seems most useful, which we have
tried to clarify and systematise here. Thus
we can think of an ARG as being any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.

The flexiblity of the gARG encoding contrasts with the classical
``event'' ARG (eARG), which, because of its origins in an idealised
mathematical model, is more limited in what can be described.
The necessity of complete precision under the eARG encoding has been
conflated with the idea of a ``true'' or ``full'' ARG.
[A discussion of how realistic inferring a completely precise ARG is]
[Do downstream applications use full ARGs? usually no]

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While the coalescent with recombination is a very useful and
robust model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
many modern datasets have properties that grossly
violate its assumptions.
For example, several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
substantially \emph{larger} than the usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with large samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Finally, the COVID-19 pandemic has provided an example of a dataset that
is clearly a very poor fit for the coalescent with recombination, but which,
nevertheless, we would like to estimate ARGs.
\cite{zhan2023towards} inferred an ARG for 1.27 million SARS-Cov-2 whole
genomes, and showed how it accurately captures several known properties of
the pathogen's evolution, and can efficiently and concisely
encode this vast dataset.

This view of ARGs,
decoupled from generative models and
without the hard requirement
of complete precision on all historical events, may help improve
methods for inference method evaluation. In the majority of cases,
ARG inference accuracy is evaluated by comparing some truth simulations
to the inferred ARGs by the pairwise comparison of local trees along the genome
using tree metrics as originally proposed by \citet{kuhner2015assessing}.
While this provides valuable insights into the method's performance,
it is not clear how well these tree distance metrics reflect
performance on real data.
In comparing tree-by-tree along the genome, the effects of recombination
structure is incorporated in a very indirect manner.
Tree distance metrics usually have $O(n^2)$ time complexity, and so cannot be
applied to the very large sample sizes currently of interest.
\citet{deng2021distribution} derived the distribution of the expected
waiting distances to tree changes.
[ \citet{ignatieva2023distribution} DID SOMETHING ELSE]
\citet{brandt2021evaluation} compared the
distributions of time to the most recent common ancestor across several
methods, which is a useful addition but only indirectly captures
topological comparisons.
% Not sure this is the right place for this sentence
All of these comparisons and evaluations have been based on
highly idealised simulations, and the effects richness of real data
(beyond simplistic error models) are almost entirely unknown.
A suite of true ARG distance metrics, that take into account more of the
global topology and can be applied to large ARGs, would be a
valuable addition to the field.
